### 2023-07-14 python_excel_ite.tex
intro pretty much done

Spreadsheet based modeling has transformed the teaching of management science in business schools over the past three decades. The late 1990's saw first editions of game changing textbooks \cite{winstonPracticalManagementScience2018,ragsdaleSpreadsheetModelingDecision2017} focused on the use of spreadsheets in teaching modeling and management science.  A series of conferences focused on teaching management science (TMS) with spreadsheets started at Dartmouth in 1998 and several presentations from that conference made up the bulk of papers in the first two issues of the INFORMS Transactions on Education journal \cite{bibid}. The next few decades saw numerous textbooks \cite{BusinessAnalyticsArt, BusinessAnalytics4th} and teaching papers \cite{INFORMSTransactionsEducation2014} that prominently featured spreadsheet based modeling.

@@

At the same time, the analytics world has been changing. The advent of programmatic analytics using tools like R and Python have changed the practice of analytics as well as the teaching landscape \cite{bibid}. Limitations of spreadsheets with respect to reproducibility and automation have certainly played a role.  Additionally, both R and Python are free, open source, and have a rich set of analytics related libraries that have contributed to their growing popularity. Python, being a general purpose programming language, also offers advantages when moving models and analytical analysis pipelines from the prototype stage into production. Enormous data science ecosystems have emerged around both R and Python providing a wealth of learning resources for both new and experienced analytics students and professionals.

Numerous R and Python based data science courses have been developed and offered at universities and through online learning platforms such as Coursera, EdX, Udemy and others. Some of these courses have made a large amount of their course content freely available for self-learners and for other educators. At our institution, we have had a spreadsheet based modeling course in place since 2001. In 2015, we created a course called Practical Computing for Data Analytics (PCDA) that introduced both R and Python (along with a little Linux) for use in data analytics work.

The PCDA course, like many introductory data science courses, focuses mostly on data wrangling, exploratory data analysis (EDA), and statistical/machine learning based predictive models. In teaching EDA using tools like ggplot2, students are challenged to think about how they would go about trying to use Excel to create some of the plots that are easy to do with R and Python based plotting libraries. There is also some limited coverage of using Python to open Excel files and do some sheet manipulation. However, other than a Python based simulation of the Monte Hall 3-door problem, there was no real attempt to do the kind of modeling that was the focus of spreadsheet based modeling textbooks.

Python has continued to make inroads to analytical territory that was formerly dominated by Excel. The widely used Python data analysis library, pandas, was created by Wes McKinney while working as a quantitative financial analyst \cite{bibid}. Python is widely used for data wrangling, visualization and predictive modeling using various flavors of regression, tree based models and neural networks. The blog \href{https://pbpython.com/}{Practical Business Python} written primarily by Chris Moffitt contributed to making Python more accessible to business analytics practitioners through high quality posts showing practical use of Python for business analysis problems. But, it got me to thinking. What about those things for which Excel is well suited such as building formula based models and doing sensitivity analysis on these models? What would those look like in Python? The importance of modeling, articulated so well in the pages of this journal \cite{powellTeachingModelingManagement2001}, has not abated. What role does Python have to play in supporting the kind of modeling taught for years in our business schools?

In 2021, I published a series of blog posts \cite{bibid} that explored this question and formed the basis of the teaching module described in this paper. During the development of these posts, I came across the \href{https://quantecon.org/}{Quant Econ: open source code for economic modeling} project which \quote{is a nonprofit organization dedicated to development and documentation of open source computational tools for economics, econometrics, and decision making}. This is a mature project, run by renowned economists \cite{bibid}, with significant support. They have developed and released numerous high quality Jupyter notebooks exploring a range of computational economic and financial modeling topics. Here was exactly the kind Python for business usage I had been seeking. With this as inspiration, I decided to turn my blog post series into a teaching module called Excel with Python (EwP).

The EwP module is not as polished nor as advanced as the Quant Econ materials, but does add to the growing body of freely available teaching materials aimed at Python for business modeling. My goals for EwP are also a little different and include:

@@

Recently there has been more recognition of the need for data scientists to learn basic software engineering principes \cite{SoftwareEngineeringData,SoftwareEngineeringDataa}.  The image of newly minted data science graduates developing spaghetti coded Jupyter notebooks not nearly ready for being put into production has reached \href{https://www.reddit.com/r/ProgrammerHumor/comments/y2xe3n/like_every_time_ever_when_the_devops_engineer/}{meme like status}. I hoped to address some of these issues in the EwP module. Sneakily, students would be exposed to some simple software engineering concepts during the process of building and analyzing familiar models. This is really no different than the integration of advanced Excel functions and techniques with spreadsheet modeling tutorials. Just as spreadsheet modeling students were able to easily wow their coworkers and managers with INDEX-MATCH and intermediate Pivot Table skills, I hoped that my students ability to properly document a Python function and use git and GitHub would give them a bit of street credibility.

>>>>

Spreadsheet based modeling has transformed the teaching of management science in business schools over the past three decades. The late 1990's saw first editions of game changing textbooks \cite{winstonPracticalManagementScience2018,ragsdaleSpreadsheetModelingDecision2017} focused on the use of spreadsheets in teaching modeling and management science.  A series of conferences focused on teaching management science (TMS) with spreadsheets started at Dartmouth in 1998 and several presentations from that conference made up the bulk of papers in the first two issues of the INFORMS Transactions on Education journal. The next few decades saw numerous textbooks \cite{BusinessAnalyticsArt, BusinessAnalytics4th} and teaching papers \cite{INFORMSTransactionsEducation2014} that prominently featured spreadsheet based modeling.

@@

At the same time, the analytics world has been changing. The advent of programmatic analytics using tools like R and Python have changed the practice of analytics as well as the teaching landscape. Limitations of spreadsheets with respect to reproducibility and automation have certainly played a role.  Additionally, both R and Python are free, open source, and have a rich set of analytics related libraries that have contributed to their growing popularity. Python, being a general purpose programming language, also offers advantages when moving models and analytical analysis pipelines from the prototype stage into production. Enormous data science ecosystems have emerged around both R and Python providing a wealth of learning resources for both new and experienced analytics students and professionals.

Numerous R and Python based data science courses have been developed and offered at universities and through online learning platforms such as Coursera, EdX, Udemy and others. Some of these courses have made a large amount of their course content freely available for self-learners and for other educators.  In 2015, we created a course called \href{http://www.sba.oakland.edu/faculty/isken/courses/mis5470/}{Practical Computing for Data Analytics} (PCDA) that introduced both R and Python (along with a little Linux) for use in data analytics work. This new course was meant as a follow up to the \href{http://www.sba.oakland.edu/faculty/isken/courses/mis5460/}{spreadsheet based modeling course} we have had in place since 2001.

The PCDA course, like many introductory data science courses, focuses mostly on data wrangling, exploratory data analysis (EDA), and statistical/machine learning based predictive models. In teaching EDA using tools like ggplot2, students are challenged to think about how they would go about trying to use Excel to create some of the plots that are easy to do with R and Python based plotting libraries. There is some limited coverage of using Python to open Excel files and do worksheet manipulation. However, other than a Python based simulation of the Monte Hall 3-door problem, there was no real attempt to do the kind of modeling that is the focus of most spreadsheet based modeling textbooks.

Python has continued to make inroads to analytical territory that was formerly dominated by spreadsheets. The widely used Python data analysis library, pandas, was created by Wes McKinney while working as a quantitative financial analyst \cite{PythonDataAnalysis}. Python is widely used for data wrangling, visualization and predictive modeling using various flavors of regression, tree based models and neural networks. The blog \href{https://pbpython.com/}{Practical Business Python} written primarily by Chris Moffitt has contributed to making Python more accessible to business analytics practitioners through high quality posts showing practical use of Python for business analysis problems. But, it got me to thinking. What about those things for which Excel is particularly well suited such as building formula based models and doing sensitivity analysis on these models? What would those look like in Python? The importance of modeling, articulated so well in the pages of this journal \cite{powellTeachingModelingManagement2001}, has not abated. What role does Python have to play in supporting the kind of modeling taught for years in our business schools?

In 2021, I published a \href{https://bitsofanalytics.org/#category=excel}{series of blog posts} that explored this question and formed the basis of the teaching module described in this paper. During the development of these posts, I came across the \href{https://quantecon.org/}{Quant Econ: open source code for economic modeling} project which \quote{is a nonprofit organization dedicated to development and documentation of open source computational tools for economics, econometrics, and decision making}. This is a mature project, run by renowned economists, with significant support. They have developed and released numerous high quality Jupyter notebooks exploring a range of computational economic and financial modeling topics. Here was an example of the Python for business usage I had been seeking. With this as inspiration, I decided to turn my blog post series into a teaching module called Excel with Python (EwP).

The EwP module is not as polished nor as advanced as the Quant Econ materials, but does add to the growing body of freely available teaching materials aimed at Python for business modeling. My goals for EwP are also a little different than those of Quant Econ and include:

@@

Recently there has been more recognition of the need for data scientists to learn basic software engineering principes \cite{SoftwareEngineeringData,SoftwareEngineeringDataa,BuildingReproducibleAnalytical}.  The image of newly minted data science graduates developing spaghetti coded Jupyter notebooks not nearly ready for being put into production has reached \href{https://www.reddit.com/r/ProgrammerHumor/comments/y2xe3n/like_every_time_ever_when_the_devops_engineer/}{meme like status}. I hoped to address some of these issues in the EwP module. Sneakily, students would be exposed to some simple software engineering concepts during the process of building and analyzing familiar models. This is really no different than the integration of advanced Excel functions and techniques with spreadsheet modeling tutorials. Just as spreadsheet modeling students were able to easily wow their coworkers and managers with INDEX-MATCH and intermediate Pivot Table skills, I hoped that my students ability to properly document a Python function and use git and GitHub would give them a bit more professional credibility.
